### 2020-08-12 manuscript/manuscript.rmd
added chain analyses

@@

# gg_n_cat

>>>>

There are several impacts of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister). It could be that there are a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categroies are available. This would suggest that we should favor the result under a very large number of categories, Porifera-sister. Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameter. This would suggest that we should favor the result under a smaller number of categories, Ctenophora-sister. One way to assess this is to look at the allocation of sites to equilibrium frequency categories. Each new category adds the same number of parameters, but that site can be allocated to any number of sites. If as we add categories those categories are allocated to a small fraction of sites, then the negative cost of those sites is high relative to the fraction of the data they help explain, and we should be skeptical of the results shen adding so many sites. That is exactly what we see (Extended Data Fig. 3.5 a-b). This suggests that support for Prorifera-sister is due to adding a very large number of parameters that explain a tiny fraction of the data.

We can further assess how, when a add parameters for a very large number of categories, the categories that are allocated to a small number of sites differ from those that are allocated to a large number of sites. It could be that the sites that are allocated to rate caetgories are very different from the sites in the common categories, in which case the added parameters are helping to explain big variation found in rare sites that models with smaller numbers of categories are unable to adequately explain. Alternatively, when a large number of caetgories are allowed it may be that categories with many sites fragment into clouds of similar categories, in which case the added parameters are soaking up small variations between sites and adding little value. We find support for this second scenario (Extended Data Fig. 3.5 d-c).

@@

summarise_sample = function( pb ){

  allocation_counts = table(pb@allocation)
  index = allocation_counts %>% names() %>% as.numeric()

  if( ! all( index %in% rownames(pb@frequencies) ) ){
    stop("Allocations are outside available category range")
  }

  # Consider only the subset of frequencies that are allocated
  frequencies = pb@frequencies[ match(index , rownames(pb@frequencies) ), ]

  # Building on https://www.statmethods.net/advstats/mds.html
  d = dist( frequencies )
  fit = cmdscale(d,eig=TRUE, k=2)

  categories = 
    tibble(
      index = index,
      count = c(allocation_counts),
      x = fit$points[,1],
      y = fit$points[,2]
    ) %>%
    arrange(desc(count)) %>%
    mutate( cumsum = cumsum(count))

  categories %<>% mutate(rank = 1:nrow(categories))

}

phil_cat_c1 = 
  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_CAT_Chain_1.chain")

phil_cat60_c1 = 
  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Philippe2009_only_choanozoa.phy_Poisson_nCAT60_Chain_1.chain")

whel_cat_c1 = 
  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT_Chain1.chain")

whel_cat60_c1 = 
  parse_phylobayes_last_sample("../trees_new/frequency/subsampled_Whelan2017_strict.phy_Poisson_CAT60_Chain_1.chain")

pb_samples = list(
  summarise_sample( phil_cat_c1 ) %>% 
    mutate ( chain=1, matrix="Philippe2009", model="Poisson+CAT"),
  summarise_sample( phil_cat60_c1 ) %>% 
    mutate ( chain=1, matrix="Philippe2009", model="Poisson+nCAT60"),
  summarise_sample( whel_cat_c1 ) %>% 
    mutate ( chain=1, matrix="Whelan2017_strict", model="Poisson+CAT"),
  summarise_sample( whel_cat60_c1 ) %>% 
    mutate ( chain=1, matrix="Whelan2017_strict", model="Poisson+nCAT60")
) %>%
  bind_rows() %>%
  mutate( analysis = paste(matrix, model, sep=" ")) %>%
  mutate( model=factor(model, levels=c("Poisson+nCAT60", "Poisson+CAT")) )

pb_samples %>% 
ggplot() + geom_line(aes(x = rank, y = cumsum, col=analysis))

gg_rank_phil = 
  pb_samples %>% 
  filter( matrix=="Philippe2009" ) %>%
  ggplot() +
  geom_line(aes(x=rank, y=count, col=model)) + 
  theme(legend.position="right")

gg_rank_whel = 
  pb_samples %>% 
  filter( matrix=="Whelan2017_strict" ) %>%
  ggplot() +
  geom_line(aes(x=rank, y=count, col=model)) + 
  theme(legend.position="right")

gg_mds_phil = 
 pb_samples %>% 
  filter( matrix=="Philippe2009" ) %>% 
  filter( model=="Poisson+CAT" ) %>% 
  ggplot() +
  geom_point(aes(x=x, y=y, size=count), alpha=0.4 ) + 
  theme(legend.position="right")

gg_mds_whel = 
 pb_samples %>% 
  filter( matrix=="Whelan2017_strict" ) %>% 
  filter( model=="Poisson+CAT" ) %>% 
  ggplot() +
  geom_point(aes(x=x, y=y, size=count), alpha=0.4 ) + 
  theme(legend.position="right")

gg_rank_phil + ggtitle("(A) Philippe2009")

gg_rank_whel + ggtitle("(B) Whelan2017")

gg_mds_phil + ggtitle("(C) Philippe2017 Poisson+CAT")

gg_mds_phil + ggtitle("(D) Whelan2017 Poisson+CAT")
